\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-15 paper.qmd
steps for hexagon binning penciled out

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, and connect the bin centroids. This process serves reduce some noise in the resulting surface shown in \pD{}.

Define a hexagonal grid across the \kD{} space, using hexagons with fixed height ($h_b$), width ($w_b$). The starting coordinates $(x_s, y_s)$ mark the lower left of the grid. <!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
Each of the \kD{} points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's hexagonal centroid ($C^{(k)}$).

@@

sc_ltr_pos <- c(0.08, 0.96)
sc_xlims <- c(-0.25, 1.25)
sc_ylims <- c(-0.25, 2.59)
nldr_scurve <- umap_scurve_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  #coord_fixed() +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)
```

@@

>>>>

With our goal being to make a representation of this 2D layout that can be lifted into high-dimensional space, the layout needs to be augmented to include neighbour information. A simple approach would be to triangulate the points and add edges. A more stable approach is to first hexagonally bin the data, and connect the bin centroids. This process serves to reduce some noisiness in the resulting surface shown in \pD{}. The steps in this process are shown in @fig-NLDR-scurve, and documented below.

@@

sc_ltr_pos <- c(0.08, 0.96)
sc_xlims <- c(-0.25, 1.25)
sc_ylims <- c(-0.25, 2.59)
nldr_scurve <- umap_scurve_scaled |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a", sc_ltr_pos)
```

@@

#### Scaling the data

It is beneficial to define the algorithm on data having a standard scale. Here we use [0, 1] scaling, but by default include the aspect ratio produced by the NLDR $(r_1, r_2, ..., r_k)$, by allowing for the upper bound to vary, $(y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})$.

When $k=2$ which is assumed for hexagon binning, $(y_{1,\text{max}}=1$ and $y_{2,\text{max}} = \frac{r_2}{r_1}\times \frac{2}{\sqrt{3}}$. (The $\frac{2}{\sqrt{3}}$ accounts for the different height and width of a regular hexagon.)

#### Hexagon binning

Define a hexagonal grid across the \kD{} space, using hexagons with fixed height ($h_b$), width ($w_b$). The starting coordinates $(x_s, y_s)$ mark the lower left of the grid. <!--We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.-->
Each of the \kD{} points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to.

#### Obtaining bin centroids

Let $h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}$ be the function that maps a point in \kD{} space to it's hexagonal centroid ($C^{(k)}$).

#### Triangulating centroids

### 2024-04-15 paper.tex
steps for hexagon binning penciled out

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, and connect the bin centroids. This process
serves reduce some noise in the resulting surface shown in \pD{}.

Define a hexagonal grid across the \kD{} space, using hexagons with
fixed height (\(h_b\)), width (\(w_b\)). The starting coordinates
\((x_s, y_s)\) mark the lower left of the grid. Each of the \kD{} points
will belong to a hexagonal bin. That is, for each
\(y \in \mathbfit{Y}\), we can (uniquely) identify the hexagon that the
point belongs to. Let
\(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be the
function that maps a point in \kD{} space to it's hexagonal centroid
(\(C^{(k)}\)).

@@

>>>>

With our goal being to make a representation of this 2D layout that can
be lifted into high-dimensional space, the layout needs to be augmented
to include neighbour information. A simple approach would be to
triangulate the points and add edges. A more stable approach is to first
hexagonally bin the data, and connect the bin centroids. This process
serves to reduce some noisiness in the resulting surface shown in \pD{}.
The steps in this process are shown in Figure~\ref{fig-NLDR-scurve}, and
documented below.

@@

\subsubsection{Scaling the data}\label{scaling-the-data}

It is beneficial to define the algorithm on data having a standard
scale. Here we use {[}0, 1{]} scaling, but by default include the aspect
ratio produced by the NLDR \((r_1, r_2, ..., r_k)\), by allowing for the
upper bound to vary,
\((y_{1,\text{max}}, y_{2,\text{max}}, ..., y_{k,\text{max}})\).

When \(k=2\) which is assumed for hexagon binning,
\((y_{1,\text{max}}=1\) and
\(y_{2,\text{max}} = \frac{r_2}{r_1}\times \frac{2}{\sqrt{3}}\). (The
\(\frac{2}{\sqrt{3}}\) accounts for the different height and width of a
regular hexagon.)

\subsubsection{Hexagon binning}\label{hexagon-binning}

Define a hexagonal grid across the \kD{} space, using hexagons with
fixed height (\(h_b\)), width (\(w_b\)). The starting coordinates
\((x_s, y_s)\) mark the lower left of the grid. Each of the \kD{} points
will belong to a hexagonal bin. That is, for each
\(y \in \mathbfit{Y}\), we can (uniquely) identify the hexagon that the
point belongs to.

\subsubsection{Obtaining bin centroids}\label{obtaining-bin-centroids}

Let \(h : \mathbb{R}^{n\times k} \rightarrow \mathbb{R}^{m\times k}\) be
the function that maps a point in \kD{} space to it's hexagonal centroid
(\(C^{(k)}\)).

\subsubsection{Triangulating centroids}\label{triangulating-centroids}
